Summer 2025 ToDo

### 2025-11-09 Relations/Paper/Conclusion.tex
paper update

In conclusion, we have formalized the basic results of ARS theory as
presented in \cite{Terese}.  Our investigations show that much of this
theory can be made effective for most languages one typically encounters
in programming language theory.
Throughout this effort, we found that the classical theorems often assume
  more than is necessary, and these distinctions become especially pronounced
  in the context of constructive logic.
  Seeking to optimize hypotheses in every proof, we obtaining marginal improvements of the standard
termination and confluence criteria.
We also identified several new ARS properties
and established their relationship with the standard notions.
This provided us with a
clearer structural understanding of how different properties interact
with one another, contributing to a more refined and
expressive formulation of ARS theory.

>>>>

In conclusion, we have formalized the basic results of ARS theory as
presented in \cite{Terese}.  Our investigations show that much of this
theory can be made effective for most languages one typically encounters
in programming language theory.
Throughout this effort, we found that the classical theorems often assume
  more than is necessary.
  % , and these distinctions become especially pronounced
  % in the context of constructive logic.
  Seeking to optimize hypotheses in every proof, we obtain marginal improvements of the standard
termination and confluence criteria.
We also identified several new ARS properties
and established their relationship with the standard notions.
This provided us with a
clearer structural understanding of how different properties interact
with one another, contributing to a more refined and
expressive formulation of ARS theory.

We also analyzed the subtleties in formulating the concept of a well-founded relation
precisely in a constructive metatheory.  We identified several classical principles
that are needed to relate these different variations of the concept.

### 2025-11-09 Relations/Paper/Formalization-pldi.tex
paper update

\subsection{\texttt{Propositions.agda}}
There are three sets of propositions given in \terese: \texttt{1.1.9}, \texttt{1.1.10}, \texttt{1.1.11}.
The formalization of the proofs are in \texttt{Propositions.agda}. These formalizations are straightforward
and as such are not the focus of this paper.

@@

\begin{proof}
    See \texttt{module Theorem-1-2-2} in \texttt{Theorems.agda}, \\
    \verb|CR→NP : R isCR → R isNP₌|\\
    \verb|NP→UN : R isNP₌ → R isUN|\\
    \verb|ii     : R isWN × R isUN → R isCR|\\
    \verb|ii+    : R isWN × R |\texttt{isUN}$^{\to}$ \verb|→ R isCR|\\
    \verb|iii    : subcommutative R → R isCR|\\
\end{proof}
\begin{proof}[Discussion.]
    In $\mathtt{UN^{→}∧WN→CR}$ a more general proof of ($ii$) is given using $\UNto$.
\end{proof}

@@

\begin{proof}
    See \texttt{module Theorem-1-2-3} in \texttt{Theorems.agda},\\
    \verb|i     : R isWN → R isUN → R isBP|\\
    $\mathtt{i^\to}$\hspace{5mm} \verb|: R isWN → R |\texttt{isUN}$^{\to}$ \verb|→ R isBP|\\
    \verb|iiSeq : R isWN → R isUN → R isRP → isWFseq- (~R R)|\\
    \verb|iii   : R isWN → R isWCR → R isRP- → dec SN → R isSN|\\
    \verb|iv    : R isCP → R isCR|
\end{proof}
\begin{proof}[Discussion.]\hfill
  \begin{description}
    \item[($i$)] In $\mathtt{i^\to}$ we again provide a slight improvement by using $\gUNto$ rather than $\gUN$.
    \item[($ii$)] This proof is inherently classical. For the closest constructive analogue to this proof we have written $\mathtt{iiSeq}$, where we replace $\gBP$ with the stronger properties $\gWN$ and $\gUN$ and we replace $\gInc$ with the weaker property $\gRP$. 
    With the original property $\gBP$ and $gRP$ we can imply the related property $gSM$ ($\gSM$ is defined in Section~\ref{sec:Hierarchy} and its relation to $gSN$ is expanded upon). The definition and further discussion of $\isWFseqm$ are provided in Section~\ref{sec:Well-foundedness}.
    \item[($iii$)] In $\mathtt{iii}$, we replace $\gInc$ with the more general property $\gRPm$. We also assume that $\gSN$ is decidable, as this allows the construction of a witness for whether an element possesses the property $\SN$, which is essential for our proof.
    \item[($iv$)] We did not formalize $\gCR \to \gCP$ as it requires cardinality assumptions. \qedhere
  \end{description}
\end{proof}

>>>>

\subsection{\texttt{Propositions.agda}}
There are three sets of propositions given in \terese: \texttt{1.1.9}, \texttt{1.1.10}, \texttt{1.1.11}.
The formalization of the proofs are in \texttt{Propositions.agda}. These formalizations are straightforward
and do not merit further discussion.

@@

\begin{proof}
    See \texttt{module Theorem-1-2-2} in \texttt{Theorems.agda}, \\
    \verb|CR→NP : R isCR → R isNP₌|\\
    \verb|NP→UN : R isNP₌ → R isUN|\\
    \verb|ii     : R isWN × R isUN → R isCR|\\
    \verb|ii+    : R isWN × R |\texttt{isUN}$^{\to}$ \verb|→ R isCR|\\
    \verb|iii    : subcommutative R → R isCR|
\end{proof}
\begin{remark}
    In $\mathtt{UN^{→}∧WN→CR}$ a more general proof of ($ii$) is given using $\UNto$.
\end{remark}

@@

\begin{proof}
    See \texttt{module Theorem-1-2-3} in \texttt{Theorems.agda},\\
    \verb|i     : R isWN → R isUN → R isBP|\\
    $\mathtt{i^\to}$\hspace{5mm} \verb|: R isWN → R |\texttt{isUN}$^{\to}$ \verb|→ R isBP|\\
    \verb|iiSeq : R isWN → R isUN → R isRP → isWFseq- (~R R)|\\
    \verb|iii   : R isWN → R isWCR → R isRP- → dec SN → R isSN|\\
    \verb|iv    : R isCP → R isCR|
\end{proof}
\begin{remark} \hfill
  \begin{description}
    \item[($i$)] In $\mathtt{i^\to}$ we again provide a slight improvement by using $\gUNto$ rather than $\gUN$.
    \item[($ii$)] This proof is inherently classical. For the closest constructive analogue to this proof we have written $\mathtt{iiSeq}$, where we replace $\gBP$ with the stronger properties $\gWN$ and $\gUN$ and we replace $\gInc$ with the weaker property $\gRP$.
    With the original property $\gBP$ and $gRP$ we can imply the related property $gSM$ ($\gSM$ is defined in Section~\ref{sec:Hierarchy} and its relation to $gSN$ is expanded upon). The definition and further discussion of $\isWFseqm$ are provided in Section~\ref{sec:Well-foundedness}.
    \item[($iii$)] In $\mathtt{iii}$, we replace $\gInc$ with the more general property $\gRPm$. We also assume that $\gSN$ is decidable, as this allows the construction of a witness for whether an element possesses the property $\SN$, which is essential for our proof.
  \end{description}
\end{remark}

### 2025-11-09 Relations/Paper/Implications.tex
wf diagram update
Minor paper changes + new WF table
paper update

Proposition \ref{prop:nftomf} suggests that we consider $\MF$ as a termination property in its own right. This leads us to
explore the following broader taxonomy of concepts relating to termination and confluence.

@@

\clearpage
\renewcommand*{\thefootnote}{\fnsymbol{footnote}}

\begin{table}[h!]
    \centering
    \renewcommand\arraystretch{1.2}
    \begin{tabular}{!{\vrule width 1.5pt}
        >{\columncolor{gray!30}}l
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c!{\vrule width 1.5pt}}
        \Xhline{1.5pt}
        \rowcolor{gray!30}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\WM$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\WN$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\SM$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\SMandWN$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\SN$} \\
        \Xhline{1.5pt}
        $\UNto$ & CE-\ref{CE:4} & CE-\ref{CE:11} & CE-\ref{CE:4} & CE-\ref{CE:6} & $\CR$\footnotemark[1] & CE-\ref{CE:8} & $\CR$\footnotemark[1] & CE-\ref{CE:4} & $\CR$\footnotemark[1] & $\SN$ \\
        \hline
        $\NP$ & CE-\ref{CE:3} & CE-\ref{CE:8} & $\CR$ & CE-\ref{CE:6} & $\CR$\footnotemark[1] & CE-\ref{CE:8} & $\CR$ & $\SN$ & $\CR$\footnotemark[2] & $\SN$ \\
        \hline
        $\MP$ & $\CR$ & CE-\ref{CE:8} & $\CR$ & CE-\ref{CE:6} & $\CR$\footnotemark[2] & CE-\ref{CE:8} & $\CR$ & $\SN$ & $\CR$\footnotemark[2] & $\SN$ \\
        \hline
        $\CR$ & $\CR$ & CE-\ref{CE:8} & $\CR$ & CE-\ref{CE:6} & $\CR$ & CE-\ref{CE:8} & $\CR$ & $\SN$ & $\CR$ & $\SN$ \\
        \hline
        \Xhline{1.5pt}
    \end{tabular}
    \caption{Local implications}
\end{table}

\vspace{-1cm}
% \\[-0.5cm]
\begin{table}[h!]
    \centering
    \renewcommand\arraystretch{1.2}
    \begin{tabular}{!{\vrule width 1.5pt}
        >{\columncolor{gray!30}}l
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c!{\vrule width 1.5pt}}
        \Xhline{1.5pt}
        \rowcolor{gray!30}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\gWM$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\gWN$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\gSM$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\gSMandWN$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\gSN$} \\
        \Xhline{1.5pt}
        $\gUNto$ & CE-\ref{CE:2} & CE-\ref{CE:8} & $\gCR$ & CE-\ref{CE:6} & $\gCR$\footnotemark[1] & CE-\ref{CE:8} & $\gCR$ & $\gSN$ & $\gCR$\footnotemark[2] & $\gSN$ \\
        \hline
        $\gNP$ & CE-\ref{CE:3} & CE-\ref{CE:8} & $\gCR$ & CE-\ref{CE:6} & $\gCR$\footnotemark[1] & CE-\ref{CE:8} & $\gCR$ & $\gSN$ & $\gCR$\footnotemark[2] & $\gSN$ \\
        \hline
        $\gMP$ & $\gCR$ & CE-\ref{CE:8} & $\gCR$ & CE-\ref{CE:6} & $\gCR$\footnotemark[2] & CE-\ref{CE:8} & $\gCR$ & $\gSN$ & $\gCR$\footnotemark[2] & $\gSN$ \\
        \hline
        $\gCR$ & $\gCR$ & CE-\ref{CE:8} & $\gCR$ & CE-\ref{CE:7} & $\gCR$ & CE-\ref{CE:8} & $\gCR$ & $\gSN$ & $\gCR$ & $\gSN$ \\
        \Xhline{1.5pt}
    \end{tabular}
    \caption{Global implications. \\$^*$~This implication also requires $\gWCR$. \\ $^\dagger$~This implication also requires either $\gWCR$ or the classical property required to go from $\SN \to \WN$ or $\SM \to \WM$.}
\end{table}
% \footnotetext[1]{This implication also requires $\gWCR$.}
% \footnotetext[2]{This implication also requires either $\gWCR$ or the classical property required to go from
% $\SN \to \WN$ or $\SM \to \WM$.}
\renewcommand*{\thefootnote}{\arabic{footnote}}

@@

\begin{remark} The following are the key takeaways from the implication tables. \hfill
    \begin{enumerate}
        \item The property $\gWCR$ can sometimes be a substitute for the classical
        property required for $\SN \implies \WN$ (and similarly $\SM \implies \WM$).
        \item $\gWN \land \gUNto \implies \gCR$ does not require any classical assumptions
         or $\gWCR$ for the implication to $\gCR$, unlike $\gSN \land \gUNto$.
        \item $\gWN \land \gUNto \implies \gCR$ only holds when the properties are global. The
        counterexample provided illustrates why $\WCR \land \WN \land \UNto \nRightarrow \CR$.
        \item We have a generalization of Newman's Lemma as $\SM \land \gWCR \implies \CR$. This is explored in Subsection \ref{subsec:newnewman}.
        \item Both $\SM$ and $\WN$ are required for $\SN$, this connection is explored in Subsection \ref{subsec:SMWNSN}.
    \end{enumerate}
\end{remark}

@@

>>>>

% BP and RP imply SM. So BP and RP and WN classically implies SN.
Proposition \ref{prop:nftomf} suggests that we consider $\MF$ as a termination property in its own right. This leads us to
explore the following broader taxonomy of concepts relating to termination and confluence.

@@

% \clearpage
\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
\begin{table}[h!]
    \centering
    \renewcommand\arraystretch{1.2}
    \begin{tabular}{!{\vrule width 1.5pt}
        >{\columncolor{gray!30}}l
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c!{\vrule width 1.5pt}}
        \Xhline{1.5pt}
        \rowcolor{gray!30}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\WM$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\WN$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\SM$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\SMandWN$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\SN$} \\
        \Xhline{1.5pt}
        $\UNto$ & CE-\ref{CE:4} & CE-\ref{CE:11} & CE-\ref{CE:4} & CE-\ref{CE:6} & $\CR$\footnotemark[1] & CE-\ref{CE:8} & $\CR$\footnotemark[1] & CE-\ref{CE:4} & $\CR$\footnotemark[1] & $\SN$ \\
        \hline
        $\NP$ & CE-\ref{CE:3} & CE-\ref{CE:8} & $\CR$ & CE-\ref{CE:6} & $\CR$\footnotemark[1] & CE-\ref{CE:8} & $\CR$ & $\SN$ & $\CR$\footnotemark[2] & $\SN$ \\
        \hline
        $\MP$ & $\CR$ & CE-\ref{CE:8} & $\CR$ & CE-\ref{CE:6} & $\CR$\footnotemark[2] & CE-\ref{CE:8} & $\CR$ & $\SN$ & $\CR$\footnotemark[2] & $\SN$ \\
        \hline
        $\CR$ & $\CR$ & CE-\ref{CE:8} & $\CR$ & CE-\ref{CE:6} & $\CR$ & CE-\ref{CE:8} & $\CR$ & $\SN$ & $\CR$ & $\SN$ \\
        \hline
        \Xhline{1.5pt}
    \end{tabular}
    \caption{Local implications}
\end{table}

\vspace{-1cm}
% \\[-0.5cm]
\begin{table}[h!]
    \centering
    \renewcommand\arraystretch{1.2}
    \begin{tabular}{!{\vrule width 1.5pt}
        >{\columncolor{gray!30}}l
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c
        !{\vrule width 1.5pt}c|c!{\vrule width 1.5pt}}
        \Xhline{1.5pt}
        \rowcolor{gray!30}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\gWM$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\gWN$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\gSM$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\gSMandWN$}
        & \multicolumn{2}{c!{\vrule width 1.5pt}}{$\gSN$} \\
        \Xhline{1.5pt}
        $\gUNto$ & CE-\ref{CE:2} & CE-\ref{CE:8} & $\gCR$ & CE-\ref{CE:6} & $\gCR$\footnotemark[1] & CE-\ref{CE:8} & $\gCR$ & $\gSN$ & $\gCR$\footnotemark[2] & $\gSN$ \\
        \hline
        $\gNP$ & CE-\ref{CE:3} & CE-\ref{CE:8} & $\gCR$ & CE-\ref{CE:6} & $\gCR$\footnotemark[1] & CE-\ref{CE:8} & $\gCR$ & $\gSN$ & $\gCR$\footnotemark[2] & $\gSN$ \\
        \hline
        $\gMP$ & $\gCR$ & CE-\ref{CE:8} & $\gCR$ & CE-\ref{CE:6} & $\gCR$\footnotemark[2] & CE-\ref{CE:8} & $\gCR$ & $\gSN$ & $\gCR$\footnotemark[2] & $\gSN$ \\
        \hline
        $\gCR$ & $\gCR$ & CE-\ref{CE:8} & $\gCR$ & CE-\ref{CE:7} & $\gCR$ & CE-\ref{CE:8} & $\gCR$ & $\gSN$ & $\gCR$ & $\gSN$ \\
        \Xhline{1.5pt}
    \end{tabular}
    \caption{Global implications. \\$^*$~This implication also requires $\gWCR$. \\ $^\dagger$~This requires either $\gWCR$ or the classical implication $\SN \to \WN$ or $\SM \to \WM$.}
\end{table}
% \footnotetext[1]{This implication also requires $\gWCR$.}
% \footnotetext[2]{This implication also requires either $\gWCR$ or the classical property required to go from
% $\SN \to \WN$ or $\SM \to \WM$.}
\renewcommand*{\thefootnote}{\arabic{footnote}}

@@

\begin{remark} The following are the key takeaways from the implication tables. \hfill
    \begin{enumerate}
        \item The property $\gWCR$ can sometimes be a substitute for the classical
        property required for $\SN \implies \WN$ (and similarly $\SM \implies \WM$).
        \item $\gWN \land \gUNto \implies \gCR$ does not require any classical assumptions
         or $\gWCR$ for the implication to $\gCR$, unlike $\gSN \land \gUNto$.
        \item $\gWN \land \gUNto \implies \gCR$ only holds when the properties are global. Counterexample~\ref{CE:4} shows that $\WCR \land \WN \land \UNto \nRightarrow \CR$.
        \item We have a generalization of Newman's Lemma as $\SM \land \gWCR \implies \CR$. This is explored in Subsection \ref{subsec:newnewman}.
        \item Both $\SM$ and $\WN$ are required for $\SN$, this connection is explored in Subsection \ref{subsec:SMWNSN}.
    \end{enumerate}
\end{remark}

@@

\subsection{A sufficient condition for $\gSN \to \gWN$}
\label{sec:SNtoWN}
It is perhaps surprising that the intuitively obvious implication from strong
normalization to weak normalization is actually not provable.
A counterexample is furnished in the file \texttt{WFCounters.agda}.
% This fact is related to the subtleties arising in constructive formulation of
% well-foundedness, as we discuss at the end of Section \ref{sec:Well-foundedness}.
At the same time, this implication is valid for all relations which are
1) decidable, and 2) finitely branching.
This is formalized in file \apcomment{Need to fill this in.}

Since these conditions are in fact satisfied by all of the standard
term rewrite systems and lambda calculi, we see that for such ARSs the implication
is automatically valid.

### 2025-11-09 Relations/Paper/Well-foundedness.tex
paper update

One may wonder whether the difference between the various definitions above matters for 
rewriting theory.  In particular, how often are the properties listed in 
Figure \ref{tab:cprop} actually valid for a given rewrite system, so that its termination 
($\WFacc$) could be inferred from a weaker statement, such as $\WFminDNEm$?

On the one hand, \cite{Berardi} proves a general result to the effect that
$\accDNE$ is valid \emph{metatheoretically} for a large class of definable relations.
Specifically, if $R$ is a primitive recursive predicate on the natural numbers and
$\WFaccm(R)$ can be proved in higher-order arithmetic, then an intuitionistic proof 
of $\WFacc(R)$ is also possible.
% situation of separating these notions is complicated by the fact that, for all
% \emph{definable} predicates, the classical and constructive notions do coincide.
% (We have not proved this, but see \cite{Berardi} for a related result.)
So there is little hope of distinguishing between these notions via an explicit counterexample.

@@

Such distinctions have consequences for rewriting theory.
For example, recall that $x\in \SN_R$ iff $x$ is $\bar{R}$-accessible, where $\bar{R}$ 
is the converse of $R$.  Thus, if $\bar{R}$ is well-founded, $R \models \gSN$.
Yet actually computing the normal form of a given $x$ seems to require a
strong form of decidability of being $R$-minimal, namely:
\[
\tag{$\isMinDec$} \forall x. \left(\exists y. Ryx\right) \lor \left(\forall y. \lnot Ryx\right)
\]
This is weaker than deciding whether $x$ is a normal form:
% Note that deciding whether $x$ is a normal form constitutes the weaker disjunction
$\lnot\lnot\left(\exists y. Ryx\right) \lor \left(\forall y. \lnot Ryx\right)$.

@@

Combined with the natural numbers example in \texttt{WFCounters.agda}, this shows
that the intuitive implication $\SN_R \to \WN_R$ is not actually intuitionistically valid.

>>>>

One may wonder whether the differences between the various definitions above matter for
rewriting theory.  In particular, how often are the properties listed in
Figure \ref{tab:cprop} actually valid for a given rewrite system, so that its termination
($\WFacc$) could be inferred from a weaker statement, such as $\WFminDNEm$?

On the one hand, \cite{Berardi} proves a general result to the effect that
$\accDNE$ is valid \emph{metatheoretically} for a large class of definable relations.
Specifically, if $R$ is a primitive recursive predicate on the natural numbers and
$\WFaccm(R)$ can be proved in higher-order arithmetic, then an intuitionistic proof
of $\WFacc(R)$ in higher arithmetic must also exist.
% situation of separating these notions is complicated by the fact that, for all
% \emph{definable} predicates, the classical and constructive notions do coincide.
% (We have not proved this, but see \cite{Berardi} for a related result.)
So there is little hope of distinguishing between these notions via an explicit counterexample.

@@

Such distinctions have consequences for rewriting theory.
For example, recall that $x\in \SN_R$ iff $x$ is $\bar{R}$-accessible, where $\bar{R}$
is the converse of $R$.  Thus, if $\bar{R}$ is well-founded, $R \models \gSN$.
Yet actually computing the normal form of a given $x$ seems to require a
strong form of decidability of being $R$-minimal, namely:
\[
\tag{$\isMinDec$} \forall x. \left(\exists y. Ryx\right) \lor \left(\forall y. \lnot Ryx\right)
\]
This is stronger than deciding whether $x$ is a normal form:
% Note that deciding whether $x$ is a normal form constitutes the weaker disjunction
$\lnot\lnot\left(\exists y. Ryx\right) \lor \left(\forall y. \lnot Ryx\right)$.
Indeed, in the file \texttt{ARS/Examples.agda} the implication $\gSN \to \gWN$ 
is equivalent to the principle of excluded middle.

@@

It follows that there can be no constructive proof of $\WFmin$ from $\WFacc$,
as the axiom of excluded middle would be a consequence of such a proof.
